---
title: "1st round of competitions: points"
output: html_document
---

```{r startup, include = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = T, eval=T, message=F, warning=F, error=F, comment=NA, cache=F, R.options=list(width=220))
```


```{r, echo=FALSE}
library(dplyr)
library(readr)
who <- read_csv("/Users/aldosolari/Documents/mygithub/DM/docs/competitions/DMcompetitions.csv")
who2 = who %>% 
  rename(Team = "Team name (repeat Username if single)", TeamN = "Number of components of the team", User = "Username (BeeViva)", Badge = "Badge number")  %>% 
  arrange(Team) %>%
  select(Surname, Name, Team, User, Badge) 
who2 = who2[-c(1,38,23,33,40,48),]
who2[who2$Surname=="Missineo","Team"] = "Derek"
who2[who2$Surname=="Pirotta","Team"] = "Derek"

who2$User[9]<- "beatrice.somaschini21"
who2$User[23]<- "giabellianna"
who2$User[36]<-"e.benincasa1"
who2$User[37]<-"f.melograna1"
who2$User[44]<- "sonia_cucchi"
who2$User[45]<-"joana.curri"

#======================
library(rvest)
library(dplyr)
url <- 'http://www.bee-viva.com/competitions/bike_sharing'
webpage <- read_html(url)

tab = webpage %>%
  html_nodes("table") %>%
  .[3] %>%
  html_table(fill = TRUE)

score_text = tab[[1]]$Punteggio
twodig = regmatches(score_text, gregexpr("[[:digit:]]+", score_text))
score = as.numeric(unlist(lapply(twodig, function(x) paste(x,collapse="."))))

max = min(score)
min = max(score)
percent = (score-min)/(max-min)*100
points = round(5*percent/100,2)
results = data.frame(User=tab[[1]]$Nome, points)
ix = match(results$User, who2$User)
pub_na = data.frame(Badge=who2$Badge[ix], Surname=who2$Surname[ix], Name=who2$Name[ix], results)
pub = pub_na[complete.cases(pub_na),]
pub_bike = pub %>% arrange(Badge)
#=========================
url <- 'http://www.bee-viva.com/competitions/ok_cupid'
webpage <- read_html(url)

tab = webpage %>%
  html_nodes("table") %>%
  .[3] %>%
  html_table(fill = TRUE)

score_text = tab[[1]]$Punteggio
twodig = regmatches(score_text, gregexpr("[[:digit:]]+", score_text))
score = as.numeric(unlist(lapply(twodig, function(x) paste(x,collapse="."))))

max = max(score)
min = min(score)
percent = (score-min)/(max-min)*100
points = round(5*percent/100,2)
results = data.frame(User=tab[[1]]$Nome, points)
ix = match(results$User, who2$User)
pub_na = data.frame(Badge=who2$Badge[ix], Surname=who2$Surname[ix], Name=who2$Name[ix], results)
pub = pub_na[complete.cases(pub_na),]
pub_okcupid = pub %>% arrange(Badge)
#============================
url <- 'http://www.bee-viva.com/competitions/ames'
webpage <- read_html(url)

tab = webpage %>%
  html_nodes("table") %>%
  .[3] %>%
  html_table(fill = TRUE)

score_text = tab[[1]]$Punteggio
twodig = regmatches(score_text, gregexpr("[[:digit:]]+", score_text))
score = as.numeric(unlist(lapply(twodig, function(x) paste(x,collapse="."))))

fdr = ecdf(score)
percent = (1-fdr(score-1))*100
points = round(5*percent/100,2)
results = data.frame(User=tab[[1]]$Nome, points)
ix = match(results$User, who2$User)
pub_na = data.frame(Badge=who2$Badge[ix], Surname=who2$Surname[ix], Name=who2$Name[ix], results)
pub = pub_na[complete.cases(pub_na),]
pub_ames = pub %>% arrange(Badge)
#============================



pub_all = data.frame(pub_ames[,1:4] ,Ames = pub_ames[,5], BikeSharing=pub_bike[,5], OkCupid=pub_okcupid[,5], Points = pub_ames[,5] + pub_bike[,5] + pub_okcupid[,5])
row.names(pub_all) = as.character(1:nrow(pub_all))
knitr::kable(pub_all, row.names = T)
```

# Adjustments

* Peer rating: 
    - Daniele Parimbelli + 0.463
    - Ongaro	Andrea + 0.451
    - Arguirov	Kristofer + 0.230
    - Perrini	Danilo + 0.452
    - Panizza	Susanna	+ 0.340 
    - Buzzini	Roberto + 0.238
    
* Non-reproducible code: points = min(submitted, published)
    - ConcoBeach (Bikesharing): min(4.57, 4.62) = 4.57
    - Blur (OkCupid): min(4.92, 4.92) = 4.92
    - Maria Letizia Sciartilli (Ames): t.b.d.
    
* Publication on github
    - Paolo D’Angelo -0.5 (All: missing Summary, Models used, Non-standard R packages, R code with unnecessary printing, but lesson learnt included)
    - Nicolò Ghioldi -0.5 (All: missing Summary)
    - Elisa Pirotta -0.5 (House prices and OkCupid: missing Summary, Non-standard R packages; Bikesharing: missing Summary)
    - Sonia Cucchi -0.5 (House prices: R code with a lot of unnecessary printing; OKCupid: missing Summary, R code with unnecessary printing)
    - Federico Melograna -0.5 (House prices: missing Summary, R code with a lot of unnecessary printing)
    - Piovesana Dario -0.5 (House prices: R code with a lot of unnecessary printing)
    - Sara Pisaniello -0.1 (OKCupid: missing Summary, R code with unnecessary printing)
    - Nicholas Giuseppe Missineo -0.1 (House prices: R code with unnecessary printing)
    - Giacomo Floriani -0.1 (House prices: missing Non-standard R packages)