---
title: "House prices"
author: Aldo Solari
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightLanguage: R
      countIncrementalSlides: false
      highlightLines: true   
---

```{r startup, include = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = T, eval=T, message=F, warning=F, error=F, comment=NA, cache=T, R.options=list(width=220))
```

# Outline

* Description
* Data
* Evaluation
* Benchmark
* Timeline


---

# House Prices

This competition uses the Ames IA housing data. There are 2930 properties in the data. The sale price was recorded along with 81 predictors, including:

* Location (e.g. neighborhood) and lot information.
* House components (garage, fireplace, pool, porch, etc.).
* General assessments such as overall quality and condition.
* Number of bedrooms, baths, and so on. 

The training set has $n=1460$, the test set has $m=1470$.

More details can be found in [De Cock (2011, Journal of Statistics Education)](http://ww2.amstat.org/publications/jse/v19n3/decock.pdf).

See also the Kaggle competition [House Prices: Advanced Regression Techniques](https://www.kaggle.com/c/house-prices-advanced-regression-techniques)

---

# Map (stolen from Max Kuhn)

```{r ames-map, echo = FALSE, message = FALSE, fig.align='center', dev = "svg"}
library(dplyr)
library(AmesHousing)
library(leaflet)
library(htmltools)
library(Cairo)
ames <- make_ames()

col_key <- c(
  'NAmes',     '#0000FF',
  'CollgCr',   '#FF0000',
  'OldTown',   '#FFFFFF',
  'Edwards',   '#FF00B6',
  'Somerst',   '#FF3030',
  'NridgHt',   '#009FFF',
  'Gilbert',   '#DD00FF',
  'Sawyer',    '#9A4D42',
  'NWAmes',    '#00FFBE',
  'SawyerW',   '#1F9698',
  'Mitchel',   '#FFACFD',
  'BrkSide',   '#720055',
  'Crawfor',   '#F1085C',
  'IDOTRR',    '#FE8F42',
  'Timber',    '#004CFF',
  'NoRidge',   '#ffff00',
  'StoneBr',   '#B1CC71',
  'SWISU',     '#02AD24',
  'ClearCr',   '#FFD300',
  'MeadowV',   '#886C00',
  'BrDale',    '#FFB79F',
  'Blmngtn',   '#858567',
  'Veenker',   '#A10300',
  'NPkVill',   '#00479E',
  'Blueste',   '#DC5E93',
  'Greens',    '#93D4FF',
  'GreenHills', '#e5f2e5', 
  'Landmrk',   '#C8FF00'
) 
col_key <- as.data.frame(matrix(col_key, byrow = TRUE, ncol = 2))
names(col_key) <- c("Neighborhood", "color")
col_key <- col_key %>%
    mutate(
      Neighborhood =
        dplyr::recode(
          Neighborhood,
          "Blmngtn" = "Bloomington_Heights",
          "Bluestem" = "Bluestem",
          "BrDale" = "Briardale",
          "BrkSide" = "Brookside",
          "ClearCr" = "Clear_Creek",
          "CollgCr" = "College_Creek",
          "Crawfor" = "Crawford",
          "Edwards" = "Edwards",
          "Gilbert" = "Gilbert",
          "Greens" = "Greens",
          "GreenHills" = "Green_Hills",
          "IDOTRR" = "Iowa_DOT_and_Rail_Road",
          "Landmrk" = "Landmark",
          "MeadowV" = "Meadow_Village",
          "Mitchel" = "Mitchell",
          "NAmes" = "North_Ames",
          "NoRidge" = "Northridge",
          "NPkVill" = "Northpark_Villa",
          "NridgHt" = "Northridge_Heights",
          "NWAmes" = "Northwest_Ames",
          "OldTown" = "Old_Town",
          "SWISU" = "South_and_West_of_Iowa_State_University",
          "Sawyer" = "Sawyer",
          "SawyerW" = "Sawyer_West",
          "Somerst" = "Somerset",
          "StoneBr" = "Stone_Brook",
          "Timber" = "Timberland",
          "Veenker" = "Veenker"
        ))

lon_rnd <- range(ames$Longitude)
lat_rnd <- range(ames$Latitude)

ia_map <- leaflet(width = "80%") %>%
  addProviderTiles(providers$CartoDB.DarkMatterNoLabels)

for(i in 1:nrow(col_key)) {
  ia_map <- ia_map %>%
    addCircles(
      data = subset(ames, Neighborhood == col_key$Neighborhood[i]),
      lng = ~Longitude, lat = ~Latitude,
      color = col_key$color[i],
      fill = TRUE,
      fillColor = col_key$color[i],
      radius = 6,
      popup = htmlEscape(col_key$Neighborhood[i]),
      opacity = .25)
}
ia_map
```

---

# Data

The training data has 82 columns which include 23 nominal, 23 ordinal, 14 discrete, and 20 continuous variables (and 2 additional observation identifiers). Here’s a brief version of what you’ll find in the data description file.

```{r, echo=FALSE}
knitr::kable(head(ames[,rev(1:ncol(ames))]), format = 'html')
```

[Data documentation](http://bee-fore.s3-eu-west-1.amazonaws.com/datasets/62.txt)

---

# Evaluation

* Submissions are evaluated on __Root-Mean-Squared-Error__ (RMSE) between the logarithm of the predicted value and the logarithm of the observed sales price. 

$$\mathrm{RMSE}= \sqrt{\frac{1}{m}\sum_{i=1}^{m}\left[\log(y^*_i)- \log(\hat{y}_i)\right]^2}$$


* Taking logs means that errors in predicting expensive houses and cheap houses will affect the result equally

* During the competition, the leaderboard displays your partial score, which is the RMSE for $m=735$ (random) houses of the test set

* At the end of the contest, the leaderboard will display the final score, which is the RMSE for the remaining $m=735$ houses of the test set

* The final score will determine the final winner. This method prevents users from overfitting to the leaderboard

* The __points__ will be calculated on the basis of the distribution of the competition' scores

---

# Benchmark


```{r, eval=F}
train <- read.csv("60.csv", stringsAsFactors=T)
test <- read.csv("61.csv", stringsAsFactors=F)

fit <- lm(log10(SalePrice) ~ Year.Built + Lot.Area + Gr.Liv.Area, data = train)
yhat = 10^predict(fit, newdata=test)

write.table(file="mySubmission.txt", yhat, row.names = FALSE, col.names = FALSE)
```

---

# Timeline

Entry: October 8

Final submission: November 16

The R code used to get the predictions __must be published online__ by November 17. 
Keep only the necessary code (no plots, numeric summaries etc. )





