---
title: "Data Mining"
output:
  pdf_document: default
  html_document: default
---

Name: 

Surname: 

Badge number: 

**Exercize I (ISLR, Chapter 5, Applied Exercize 8)**

We will now perform cross-validation on a simulated data set.

a. Generate a simulated data set as follows:
  
```{r}
set.seed(1)
x = rnorm(100)
y = x - 2*x^2 + rnorm(100)
```

In this data set, what is $n$ and what is $p$? Write out the model
used to generate the data in equation form.

*Write here your answers*.

b. Create a scatterplot of $X$ against $Y$. Comment on what you find.

```{r}
# write here the R code
```

*Write here your comments*.

c. Set a random seed, and then compute the LOOCV errors that
result from fitting the following four models using least squares:

(i). $Y = \beta_0 + \beta_1 X + \epsilon$

(ii). $Y = \beta_0 + \beta_1 X + \beta_2 X^2 + \epsilon$

(iii). $Y = \beta_0 + \beta_1 X + \beta_2 X^2 + \beta_3 X^3 + \epsilon$

(iv). $Y = \beta_0 + \beta_1 X + \beta_2 X^2 + \beta_3 X^3 + \beta_4 X^4 + \epsilon$

Note you may find it helpful to use the data.frame() function
to create a single data set containing both $X$ and $Y$.

```{r}
# write here the R code
```

d. Repeat c. using another random seed, and report your results.
Are your results the same as what you got in c.? Why?

```{r}
# write here the R code
```

*Write here your answers*.

e. Which of the models in c. had the smallest LOOCV error? Is
this what you expected? Explain your answer.

```{r}
# write here the R code
```

*Write here your answers*.

