---
title: "Data Mining"
subtitle: "Additive Models"
author: Aldo Solari
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightLanguage: R
      countIncrementalSlides: false
      highlightLines: true    
---

```{r startup, include = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = T, eval=T, message=F, warning=F, error=F, 
                      comment=NA, cache=F, R.options=list(width=220))
```





# Additive models

The additive model for regression is that the conditional expectation function is a sum of partial response functions, one for each predictor variable
$$\mathbb{E}(Y|X=x) = \beta_0 + \sum_{j=1}^{p}f_j(x_j)$$

This includes the linear model as a special case, where
$$f_j(x_j) = \beta_j x_j$$
but it's
clearly more general, because the $f_j$'s can be arbitrary nonlinear functions

The idea is still that each input feature makes a separate contribution to the response,
and these just add up (hence "partial response function"), but these contributions don’t have to be strictly proportional to the inputs

We do need to add a
restriction to make it identifiable; without loss of generality
$$\mathbb{E}(Y) = \beta_0, \qquad \mathbb{E}(f_j(X_j)) =0$$

How can we fit additive models?

---

Linear models assume
$$\mathbb{E}(Y|X=x) = \sum_{j=0}^{p} \beta_j x_j$$
where $x_0$ is always the constant 1

Suppose we don't condition on all $X=(X_0,\ldots,X_p)'$ but just one component of it, say $X_k$. Then 
$$\mathbb{E}(Y|X_k=x_k) =  \beta_k x_k + \mathbb{E}(\sum_{j\neq k} \beta_j X_j|X_k=x_k)$$

Then
$$\mathbb{E}(Y^{(k)}|X_k=x_k) =  \beta_k x_k$$
where $Y^{(k)} = Y - \sum_{j\neq k} \beta_j X_j$ the $k$th __partial residual__: the difference between $Y$ and what we expect it to be __ignoring__ the contribution from $X_k$

This suggests the following estimation scheme for linear models, known as the
__Gauss-Seidel algorithm__, or more commonly as __back-fitting__

---

* Given: 
    - $n\times p$ matrix $\mathbf{X}$ of $p$ predictors
    - $n\times 1$ response vector $\mathbf{y}$
    - small tolerance $\delta > 0$

* Center $\mathbf{y}$ and each column of $\mathbf{X}$

* Initialize $\hat{\beta}_{j} \leftarrow 0$ for $j=1,\ldots,p$

* Until (all $|\hat{\beta}_j - \gamma_j| \leq \delta$ ){

    for $k=1,\ldots, p$ {
    
      $y_i^{(k)} = y_i - \sum_{j\neq k} \hat{\beta}_jx_{ij}$
    
      $\gamma_k \leftarrow$ regression coefficient of $y^{(k)}$ on $x_k$
    
      $\hat{\beta}_j \leftarrow  \gamma_k$ }}

* $\hat{\beta}_{0} \leftarrow \bar{y} - \sum_{j=1}^{p} \hat{\beta}_j \bar{x}_{j}$ with original data

* Return $\hat{\beta}_0, \hat{\beta}_1, \ldots, \hat{\beta}_p$


---

Generate a response $Y$ and predictors $X_1$, $X_2$ and $X_3$ with $n = 100$ as follows:

```{r}
n = 100
set.seed(1)
x1 = rnorm(n)
x2 = rnorm(n)
x3 = rnorm(n)
y = 30 - 10*x1 + 20*x2 + 30*x3 + rnorm(n)
fit = lm(y ~ x1+x2+x3)
X = model.matrix(fit)
```

---

```{r}
# center y and X
yc = y - mean(y)
Xc = apply(X[,-1],2,function(x) x-mean(x))
# initialize 
p = ncol(Xc)
hatbeta = rep(0,p)
delta = 0.1
gamma = rep(2*delta,p)
while ( !all( abs(gamma - hatbeta) <= delta) ){
for (k in 1:p){
  gamma = hatbeta
  yk = yc - Xc[,-k, drop=F] %*% hatbeta[-k, drop=F]
  gammak = coef(lm(yk ~ 0+ Xc[,k]))
  hatbeta[k] <- gammak
}}
hatbeta0<- mean(y) - sum(hatbeta*apply(X[,-1],2,mean))
data.frame( backfitting=c(hatbeta0,hatbeta), leastsquares=coef(fit))
```

---

# Backfitting additive models

* Define the partial residuals by analogy with the linear case, as
$$Y^{(k)} = Y - \left(\beta_0 - \sum_{j\neq k} f_j(x_j) \right)$$

* Then 
$$\mathbb{E}(Y^{(k)}|X_k=x_k) =  f_k(x_k)$$

* We could use nearest neighbors, or splines, or kernels, or local-linear regression, or anything else

---

* Given: 
    - $n\times p$ matrix of $p$ predictors 
    - $n\times 1$ response vector 
    - maxit: maximum number of iterations
    - one-dimensional smoother $s$ 

* Initialize $\hat{\beta}_0 \leftarrow \bar{y}$ and $\hat{f}_{j} \leftarrow 0$ for $j=1,\ldots,p$

* for (i in 1:maxit){

    for $k=1,\ldots, p$ {
    
      $y_i^{(k)} = y_i - \sum_{j\neq k} \hat{f}_j(x_{ij})$
    
      $g_k \leftarrow s( y^{(k)} \sim x_k)$
    
      $g_k \leftarrow g_k - n^{-1} \sum_{i=1}^{n}g_k(x_{ik})$
    
      $\hat{f}_{k} \leftarrow g_k$
      
      }}

* Return $\hat{\beta}_0, \hat{f}_{1}, \ldots, \hat{f}_{p}$

---

```{r}
am_backfit <-
function(X, y, maxit=10L)
{
  p <- ncol(X)
  id <- seq_len(nrow(X))
  alpha <- mean(y)
  f <- matrix(0, ncol = p, nrow = nrow(X))
  models <- vector("list", p + 1L)
  for (i in seq_len(maxit))
  {
    for (j in seq_len(p))
    {
    p_resid <- y - alpha - apply(f[, -j], 1L, sum)
    id <- order(X[,j])
    models[[j]] <- smooth.spline(X[id,j], p_resid[id])
    f[,j] <- predict(models[[j]], X[,j])$y
    }
  alpha <- mean(y - apply(f, 1L, sum))
  }
  models[[p + 1L]] <- alpha
return(models)
}
```

---

```{r}
am_predict <-
function(models, X_new)
{
p <- ncol(X_new)
f <- matrix(0, ncol = p, nrow = nrow(X_new))
for (j in seq_len(p))
{
f[,j] <- predict(models[[j]], X_new[,j])$y
}
y <- apply(f, 1L, sum) + models[[p + 1L]]
list(y=y, f=f)
}
```

---

```{r}
set.seed(123)
n <- 500; p <- 4
X <- matrix(runif(n * p, min = -2, max = 2), ncol = p)
f1 <- cos(X[,1] * 4) + sin(X[,1] * 10) + X[,1]^(2)
f2 <- -1.5 * X[,2]^2 + (X[,2] > 1) * (X[,2]^3 - 1)
f3 <- 0
f4 <- sign(X[,4]) * 1.5
f1 <- f1 - mean(f1); f2 <- f2 - mean(f2)
f3 <- f3 - mean(f3); f4 <- f4 - mean(f4)
y <- 10 + f1 + f2 + f3 + f4 + rnorm(n, sd = 1.2)
```


---

```{r}
models <- am_backfit(X, y, maxit = 1)
yhat <- am_predict(models,X)
plot(X[,1],yhat$f[,1], type="p")
points(X[,1], f1, col=2)
```


---

# mgcv

The `mgcv` R package  is based not on
backfitting, but rather on something called the *Lanczos
algorithm*, a way of efficiently calculating truncated matrix
decompositions that is beyond the scope of this course

The basic syntax is

```{r, eval=FALSE}
fit <- gam(y ~ s(x1) + s(x2), data=train)
```

One can add arguments to the `s()` function, but the default is to use a __natural cubic spline basis__ and to automatically choose the smoothing parameter $\lambda$ via optimization of the __GCV__

---

# Interaction terms

One way to think about additive models, and about (possibly) including interaction terms, is to imagine doing a sort of __Taylor series expansion__ of the true regression function

The zero-th order expansion would be a constant:
$$f(x)\approx \beta_0$$

A purely additive model would correspond to a first-order expansion:
$$f(x)\approx \beta_0 + \sum_{j=1}^{p}f_j(x_j)$$

Two-way interactions come in when we go to a second-order expansion:
$$f(x)\approx \beta_0 + \sum_{j=1}^{p}f_j(x_j) + \sum_{j=1}^{p}\sum_{k=j+1}^{p}f_{jk}(x_j,x_k)$$

For identifiability, we need $\mathbb{E}(f_{jk}(X_j,X_k) ) = 0$

---

# Thin-plate spline

Suppose we have two input variables, $x$ and $z$, and a single response $y$. How could we do a spline fit?

One approach is to generalize the spline optimization problem so that we penalize the curvature of the spline surface (no longer a curve)

The appropriate penalized least-squares objective function to minimize is
$$\sum_{i=1}^{n}(y_i - f(x_i,z_i))^2 + \lambda \int \left[  \left(\frac{\partial^2 f}{\partial x^2} \right)^2 + 2\left(\frac{\partial^2 f}{\partial x \partial z} \right)^2 + \left(\frac{\partial^2 f}{\partial z^2} \right)^2 \right]$$

The solution is called a __thin-plate spline__  ( `s(x,z)` in `mgcv` )

This is appropriate when the two predictors $x$ and $z$ are measured on similar scales

---

# Tensor product spline

An alternative is use the spline basis functions 
$$f(x,z) = \sum_{j=1}^{K_x}\sum_{k=1}^{K_z} \beta_{jk} b_{j}(x) b_{k}(z)$$

Doing all possible multiplications of one set of numbers or functions with another is said to give their outer product or tensor product, so this is known as a __tensor product spline__   ( `te(x,z)` in `mgcv` )

This is appropriate when the measurement scales of $x$ and $z$ are very different

We have to chose the number of terms to include for each variable ( $K_x$ and $K_z$)

---

# Generalized additive models

* A natural step beyond generalized linear models is generalized additive models (GAMs), where instead of making the transformed mean response a linear function of the predictors, we make it an additive function of the predictors 

* For example
$$\mathrm{logit}\{ \mathbb{E}(Y|X=x) \} = \beta_0 + f_1(x_1) + f_2(x_2) + \ldots + f_p(x_p)$$

---

# Application: California house prices

* The Census Bureau divides the U.S. into geographic regions called __tracts__ of a few thousand people each

*  Data from the 2011 American Community Survey, containing information on the housing stock and economic circumstances of every tract in California

    -  __Median_house_value__ : The median value of the housing units in the tract (response)
    - __POPULATION__, __LATITUDE__, __LONGITUDE__ : The population, latitude and longitude of the tract
    - __Median_household_income__, __Mean_household_income__ : The median and mean income of households (in dollars, from all sources)
    - __Total_units__, __Vacant_units__ : The total number of units and the number of vacant units
    - __Owners__ : The percentage of households which own their home
    - __Median_rooms__ : The median number of rooms per unit
    - __Mean_household_size_owners__, __ Mean_household_size_renters__ The mean number of people per household which owns its home, the mean
number of people per renting household

---

```{r}
library(RCurl)
calif <- read.csv("https://raw.githubusercontent.com/aldosolari/DM/master/docs/lezioni/L15/calif.csv")
library(mgcv)
fit <- gam(log(Median_house_value) ~ s(Median_household_income) +
                   s(Mean_household_income) + 
                   s(POPULATION) + 
                   s(Total_units) + 
                   s(Vacant_units) +
                   s(Owners) + 
                   s(Median_rooms) + 
                   s(Mean_household_size_owners)  + 
                   s(Mean_household_size_renters) + 
                  s(LATITUDE,LONGITUDE), data=calif)
```

---

```{r}
plot(fit, se = 2, shade = TRUE, resid = TRUE, select=1)
```

---

```{r}
plot(fit, select = 10, se = FALSE)
```

---

```{r}
plot(fit, select = 10, phi = 60, scheme = TRUE, ticktype = "detailed", cex.axis = 0.5)
```


